---
title             : "The title"
shorttitle        : "Title"

author: 
  - name          : "First Author"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Ernst-August Doelle"
    affiliation   : "1,2"
    role:
      - "Writing - Review & Editing"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "Wilhelm-Wundt-University"
  - id            : "2"
    institution   : "Konstanz Business School"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
  
  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
  
  One sentence clearly stating the **general problem** being addressed by  this particular study.
  
  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
  
  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
  
  One or two sentences to put the results into a more **general context**.
  
  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "keywords"
wordcount         : "X"

bibliography      : "r-references.bib"

floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

# Introduction

Interessante come 

The use of ordinal data is widespread in Psychology. Usually items from questionnaires are created using Likert scales where a certain psychological traits is measured using an intuitive scale with 3-5 or more anchor points. These measure cannot be considered *metric* measure on interval or ratio scales [stevens] but the categories are ordered.

Other examples here

Despite the usage of ordinal variables, statistical models made for these type of data are rarely used in Psychology. @Liddell2018-wu reported that the majority of published papers using likert-like measures used standard methods to analyze the data. In practical terms, they use *metric* models where the response variable cannot be considered fully numeric (vedi se c'è un termine tipo scala a rapporti).

altro sui modelli

Theoretically, using a *metric* for ordinal data is not appropriate but understanding the actual impact is not straightforward. @Liddell2018-wu did a comprehensive work about pitfalls of analzying ordinal data as metric. They showed that metric models produce higher type-1 and type-2 errors compared to the ordinal models. In particular, given the bounded nature of ordinal data, difference in the underlying latent distribution are not always captured by the metric model that simply estimate the mean of the ordinal variable. This is even more relevant when the underluing variance of the ordinal variables are not homogeneous. Furthermore they presented some situations where the metric model could be wrong in the opposite direction, finding an effect with the wrong sign (type-s error). 

Finally, they also demonstrated that even in the best condition where e.g. comparing two groups the variance are equal, the ordinal model is more powerful than the metric model given the underestimation of the true effect size from the latter.

> Some authors have argued that, despite the ordinal character of individual Likert items, averaged ordinal items can have an emergent property of an interval scale and so it is appropriate to apply metric methods to the averaged values (e.g., Carifio & Perla, 2007, 2008).

vedi come gestire questo

anche questo è importante
> Ordered-probit models typically assume that the thresholds (θ k) are the same across all groups because the thresholds are theoretically linked to the response measure, not to the predictor value. For example, when asked, “How happy are you?” with response options ‘1’ = very unhappy, ‘2’ = mildly unhappy, ‘3’ = neutral, ‘4’ = mildly happy, ‘5’ = very happy,” the latent thresholds between ordinal levels are assumed to be implicit in the phrasing of the question, regardless of other aspects of the respondent or situation. In other words, the thresholds are assumed to be part of the measurement procedure, not dependent on the value of the predictor or covariate. This can be technically referred to as a type of measurement invariance.

Da citare come paper iniziale sui modelli [@McCullagh1980-cw].

questo per la tassonomia dei diversi modelli ordinali [@Tutz2022-dg]

da vedere magari per una critica tipo Gomila sul binary model [@Robitzsch2020-la]

 
\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
